Improving Haskell Theory Exploration

### 2015-11-17 TransferReport/background.tex
Updates

\iffalse
TODO{2015-11-14}

Background

2.1 Introduce language signature
 b) stuff from QuickCheck/Spec

2.2? Machine learning and feature extraction
~1 page

2.3: QuickSpec, HipSpec
\fi

\iffalse

Mathematics provides us with powerful, systematic methods of reasoning, which we can bring to bear on this challenge; in particular those of \emph{formal logic} and \emph{statistics}. By (partially) mechanising these approaches, in the fields of \emph{theorem proving} and \emph{machine learning}, respectively, we can leverage these increasing machine capabilities and direct them for the purpose of analysis. However, the question still remains: on what should we focus that analysis?

In this work, we investigate the notion of \emph{interestingness} in the exploration of formal systems (an area known as \emph{theory exploration}) as a way to make productive use of resources in an often intractable domain. To keep things concrete, we focus our formal analysis on equational formulae describing programs in the Haskell language, for reasons elaborated in \S \ref{haskell}. Conceptually, we maintain a broader view, and survey many related areas which may offer insights on the problem.

Appeals to interestingness arise when more direct measures, such as utility, are not available. For example, the inclusion of particular statements in a program or proof development can be easily justified based on their contribution to the overall solution; however, in a \emph{library} there is no particular problem being solved, in which case we must judge statements on less direct criteria, such as how ``interesting'' they may be to our users. Appeals to interestingness abound in the history of computer-assisted reasoning; for example, in 1971 Plotkin \citep{plotkin1971further} considered the task of \textquote{discovering theorems $T$ from a system of axioms $Ax$}, and in particular the questions \textquote{Under what conditions is $T$ an interesting, possible theorem in the system $Ax$?} and \textquote{Is there a way to generate (most) interesting possible theorems?}. Despite such widespread use of the term, there is no standard definition of what makes a formal object, whether it is an axiom, a conjecture, a proof, etc., ``interesting''; although many ad-hoc heuristics have been proposed.

We begin our undertaking in \S \ref{background} by introducing the Haskell language, as well as the relevant fields of verification for context. We define a formal framework for our investigation, and show how it relates to the existing theorem proving landscape. A selection of theorem proving scenarios which \emph{require} exploration are discussed in \S \ref{examples}, whilst related work, including existing defintions of interestingness, is surveyed in \S \ref{related}. We also review the use of exploration in other fields of Artificial Intelligence and Machine Learning, where researchers are experimenting with replacing \emph{explicit} goals and rewards with \emph{implicit} alternatives such as interestingness. Recent efforts in this area have lead to the emergence of principled theories, mostly based around (algorithmic) information theory, which may be adapted to our theory exploration context.

We discuss our present contributions in \S \ref{current} and future research directions in \S \ref{future}, before concluding in \S \ref{conclusion}.

\fi

@@

\begin{figure}
  \begin{equation*}
    \begin{split}
      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
                         |\ & \texttt{Lit}\ literal                                  \\
                         |\ & \texttt{App}\ expr\ expr                               \\
                         |\ & \texttt{Lam}\ \mathcal{L}\ expr                        \\
                         |\ & \texttt{Let}\ bind\ expr                               \\
                         |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
                         |\ & \texttt{Global}\ \mathcal{G}                           \\
      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )                      \\
      altcon\  \rightarrow\ & \texttt{DataAlt}\ \mathcal{G}                          \\
                         |\ & \texttt{LitAlt}\ literal                               \\
                         |\ & \texttt{Default}                                       \\
      bind\    \rightarrow\ & \texttt{NonRec}\ \mathcal{L}\ expr                     \\
                         |\ & \texttt{Rec}\ [ ( \mathcal{L},\ expr ) ]
    \end{split}
  \end{equation*}
  where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers
  \end{tabular}

@@

The full Haskell language is rather complex, so we focus instead on \emph{Core}, an intermediate representation of the Glasgow Haskell Compiler (GHC) based on \fc. For a full treatment of \fc, and its use in GHC, see \citep[Appendix C]{sulzmann2007system}. The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} we erase types and use a custom representation of names. There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.

\subsection{QuickCheck}

@@

However, this is \emph{still} not enough: since this function only uses generic numeric operations, it will be polymorphic; allowing all numeric types\footnote{Its type will be of the form \hs{forall t. Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric. There will also be extra contraints like \hs{Eq} due to the use of \hs{==} in the unit tests.}. If we provide a fractional value, the function will again diverge. Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.

Haskell offers a partial solution to this problem in the form of \emph{property checking}\footnote{See section \ref{propertychecking} for more information}. Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion. Here is how we might restate our unit tests as properties:

@@

The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with closed values. In the \hs{factorial} example, this manifests in two ways:

\begin{itemize}
  \item \qcheck{} cannot test polymorphic functions; they type must be \emph{monomorphised} first (instantiated to a particular concrete type). This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.

@@

\iffalse

Most work in mechanised, formal mathematics focuses on a \emph{theorem proving} paradigm, usually either \emph{interactive theorem proving} (ITP) or \emph{automated theorem proving} (ATP). Whilst the boundaries between these approaches can be somewhat blurred, we give precise definitions in \S \ref{itp} and \S \ref{atp}, respectively, and characterise both as being \emph{goal-driven}: particular conjectures must be provided as input to the process, and the output is either a proof, a refutation or ``don't know'' (in the case of incomplete procedures).

In this work, we instead adopt the paradigm of \emph{(automated) theory exploration}, discussed in \S \ref{theoryexploration}, which includes the ability to \emph{generate} conjectures, and hence output \emph{novel} theorems. Since many theories will admit an infinite number of trivial theorems (e.g. $\top$, $\top \land \top$, $(\top \land \top) \land \top$, \dots) we require some mechanism for avoiding such undesirable, yet provable, statements. This is the role of interestingness, which we can use to \emph{judge} a statement, in a potentially rich way, rather than merely constructing or failing to construct a proof.

\fi

\subsection{Interactive Theorem Proving}
\label{itp}

ITP is based around a \emph{proof checker} $C$, which is a decision procedure for determining if a given value $P$ (known as a \emph{proof object} or \emph{witness}) consititutes a proof of a given statement $S$:

$$ C \colon (P \times S) \rightarrow Boolean $$

The process of ITP can hence be understood as the search for an appropriate $P$ for our \emph{goal} $S$:

$$ ITP \colon (S \times C) \rightarrow \{ P \mid C(P, S) = True \} $$

It just so happens that a useful way to implement such a system is via pure functional programming, with the result that many ITP systems (AKA \emph{proof assistants}) such as Coq and Agda appear very similar to languages like Haskell. In particular, we can represent statements $S$ as types and proofs $P$ as values, in which case the proof checker $C$ is simply a type-checker. There are some obvious quirks, such as the need to be \emph{total} (not Turing-complete) in order to ensure soundness, but overall many of the features we introduced for Haskell, such as parametricity, type classes and ADTs are directly translatable to the ITP setting.

This coincidence is due to the \emph{Curry-Howard correspondence} \citep{wadler2015propositions}, which identifies programming languages with systems of logic. Functional programming languages correspond to intuitionistic logics, and are hence a natural fit for reasoning on computers. With additional axioms, we can extend these to more familiar classical logics, although these can make it harder to compute.

Most differences between functional programming and ITP stem from different \emph{expectations} of the user. In the case of Haskell, the desire for expressivity outweighs the desire for soundness, and hence its designers opted to make it Turing-complete. For an ITP language like Agda, soundness far outweighs the inconvenience of having to prove termination, so the opposite tradeoff is made. Similarly, since many ITP ``programs'' (proofs) will never be executed, the language can avoid optimisation in favour of simplicity (and hopefully correctness). This approach is summed up by the \emph{de Bruijn criterion}, which requires that the system \textquote{generates 'proof-objects' (of some form) that can be checked by an 'easy' algorithm} \cite[\S~2]{barendregt2001proof}. One nice consequence of this approach, which is followed for example by Isabelle \citep{nipkow2002isabelle} and Coq \citep{bertot2013interactive}, is that the proof assistants themselves can become arbitrarily complex and even buggy, yet as long as the proof checker remains simple we can maintain a high degree of confidence in the results.

Their emphasis on \emph{interactivity}, mostly caused by operating in undecidable domains, means ITP can require an enormous effort for non-trivial proof or verification tasks (e.g. see \citep{hales2015formal}). One common way to mitigate this problem is by implementing powerful \emph{tactics}: meta-programs which automate as much of a proof as possible. The most striking example of such meta-programming is the \emph{Sledgehammer} component of Isabelle/HOL \citep{journals/iandc/MengQP06}, which invokes a multitude of ATP systems on (a translated form of) the current goal to see if any can provide a proof which Isabelle's core checker will accept.

\subsection{Automated Theorem Proving}
\label{atp}

ATP systems are similar in principle to proof assistants, but are based around a \emph{proof search} algorithm rather than a proof checker. By using a fixed algorithm for proving, such as \emph{resolution} \cite[\S~9.6]{Russell:2003:AIM:773294} or \emph{superposition} \citep{bachmair1994rewrite}, ATP programs are limited to particular decidable or semi-decidable fragments of logic. In particular, most ATP systems (such as E \citep{schulz2013system} and Vampire \citep{riazanov2003implementing}) operate in classical first-order logic. This is the most striking difference from ITP systems (e.g. Coq, Agda and Isabelle) which operate in higher-order logics.

In addition to its interest to logicians, ATP has been actively researched in the field of artificial intelligence, dating back to the founding of the field at the 1956 Dartmouth conference. Even by that time Newell and Simon had developed their Logic Theory Machine \citep{newell1956logic}, which was subsequently able to prove theorems like those in Principia Mathematica \citep{newell1958elements}. The approaches now known as \emph{good old-fashioned AI} (GOFAI) were due in part to the success of automated theorem proving, as attempts were made to formulate many problems in a way amenable to these powerful first-order reasoners.

Recently there has been a trend away from this direction, towards statistical formulations amenable to machine learning, which we will review in \S \ref{related}.

@@

\begin{figure}
  \begin{equation*}
    \begin{split}
      name\   =\ & N_1\ |\ \dots\ |\ N_n                               \\
      type\   =\ & T_1\ |\ \dots\ |\ T_m\ |\ type \rightarrow type   \\
      sig\    =\ & \{(n_i | (name, type); sig                                     \\
      expr_{\mathcal{s}}\ =\ & v    : t\ \text{if}\ v : t \in s                        \\
              |\ & f\ x : b\ \text{if}\ f : a \in expr_s \wedge x : a \in expr_s \\
      conj_s\ =\ & \{ a =_t b \mid a : t \in expr_s \wedge b : t \in expr_s \}
    \end{split}
  \end{equation*}

  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_i$ and types $T_i$.}
  \label{coresyntax}
\end{figure}

The most striking difference between theory exploration (TE) and theorem proving is the former's lack of an explicit goal (e.g. a user-provided conjecture to prove). Instead, the \emph{implicit} goal is more open-ended: the discovery of ``interesting'' conclusions from the given premises. Of course, a key question to ask is what do we mean by ``interesting''? There is no single answer, although many approximate measures have been proposed. The choice of what counts as interesting has been used to classify various TE implementations \citep{warburtonscaling}, along with their approaches to term generation, well-formedness and proof.

Following \citep{warburtonscaling}, we call a pair $(\Sigma, V)$, of a signature and a set of variables, a \emph{theory}. \emph{Theory exploration} then refers to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing terms of the theory which are well-formed, provable and ``interesting''.

The conditions of well-formedness and provability can be handled through the use of type systems and automated theorem provers. Hence the method of conjecture generation, and the determination of what is interesting, are the most important properties of any TE system.

Early implementations like \textsc{Theorema} \citep{buchberger2000theory} provided interactive environments, similar to computer algebra systems and interactive theorem provers, to assist the user in finding theorems. Like in the ITP approach, the relatively simple process of verification is automated, whilst the more difficult tasks (term generation and the determination of their interest) are left up to the user.

Subsequent systems have investigated \emph{automated} theory exploration, for tasks such as lemma discovery \citep{Hipster}. By removing user interaction, these properties must be implemented by algorithms. In existing systems these are tightly coupled to improve efficiency, which makes it difficult to try different approaches independently.

As an example, \qspec{} \citep{QuickSpec} discovers equations about Haskell code, which are defined as ``interesting'' if they cannot be simplified using previously discovered equations. The intuition for such criteria is to avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc. when we already know $0 + x = x$. Whilst this interestingness judgement is elegantly implemented with a congruence closure relation (version 1) and a term rewriting system
(version 2), the conjecture generation is performed in a brute-force way.

Although \qspec{} only \emph{tests} its equations rather than proving them, it is still used as the exploration component of more rigorous systems like \hspec{} and \textsc{Hipster}.

\hspec{} is currently the most powerful TE system for Haskell, and also forms a major component of \textsc{Hipster}. It uses off-the-shelf ATP programs (including AltErgo, Vampire, Z3, E, SPASS and CVC4) to verify the conjectures of \qspec{}. \qspec{}, in turn, enumerates all type-correct combinations of the terms in the theory up to some depth, groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes. This approach works well as a lemma generation system, making \hspec{} a capable inductive theorem prover as well as a theory exploration system \citep{claessen2013automating}. \hspec{} is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run \hspec{} alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.

In fact, there are similarities between the way a TE system like \hspec{} can generalise from proving \emph{particular} theorems to \emph{inventing} theorems, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test. There are likely lessons that each can learn from the other's approach to term generation.

We identify the following areas for researching potential improvements in existing TE systems such as \hspec{}:

\begin{description}
\item{Term generation}:
  Enumerating all type-correct terms is a brute-force solution to this question. Scalable alternatives to brute-force algorithms are a well-studied area of Artificial Intelligence and Machine Learning \iffalse TODO{2015-11-15}: Related work reference \fi. In particular, heuristic search algorithms like those surveyed in \citep{blum2011hybrid} could be used.
  We could also use Machine Learning methods like those used for \emph{relevance filtering} \ref{relevance} to identify some sub-set of a given theory, to prioritise over the rest.

\item{Interestingness}
  Various alternative interestingness criteria have been proposed, which we survey in \S \ref{relatedwork}. Augmenting or replacing the criteria may be useful, for example to distinguish useful relationships from incidental coincidences; or to prevent surprising, insightful equations from being discarded because they can be simplified.
\end{description}

% Theory exploration is similar to \emph{experimental mathematics}

% - Relation to Science
%  - Testable/falsifiable hypotheses are like evaluable terms (or, more generally, conjectures which can be decided, using a reasonable amount of resources).
% - Relation to AI tasks: exploring surroundings, etc.

% - Statistics is another area that's less straightforward than normal numerical computing, since there is subjectivity and judgement involved in the answering of questions.

% Theory formation: Alison? Others.
% Theory exploration: Buchberger, Moa in Isabelle, Koen in Haskell. Others?
% Theorem proving: Well-trodden: first-order ATP, higher-order ITP, functional programming
% Communication: Latex, Wikis, APIs, communicating with aliens

>>>>

@@

\begin{figure}
  \begin{equation*}
    \begin{split}
      expr\    \rightarrow\ & \texttt{Var}\ id                                       \\
                         |\ & \texttt{Lit}\ literal                                  \\
                         |\ & \texttt{App}\ expr\ expr                               \\
                         |\ & \texttt{Lam}\ \mathcal{L}\ expr                        \\
                         |\ & \texttt{Let}\ bind\ expr                               \\
                         |\ & \texttt{Case}\ expr\ \mathcal{L}\ \left[ alt \right]   \\
      id\      \rightarrow\ & \texttt{Local}\ \mathcal{L}                            \\
                         |\ & \texttt{Global}\ \mathcal{G}                           \\
      literal\ \rightarrow\ & \texttt{LitNum}\ \mathcal{N}                           \\
                         |\ & \texttt{LitStr}\ \mathcal{S}                           \\
      alt\     \rightarrow\ & ( altcon,\ [\mathcal{L}],\ expr )                      \\
      altcon\  \rightarrow\ & \texttt{DataAlt}\ \mathcal{G}                          \\
                         |\ & \texttt{LitAlt}\ literal                               \\
                         |\ & \texttt{Default}                                       \\
      bind\    \rightarrow\ & \texttt{NonRec}\ \mathcal{L}\ expr                     \\
                         |\ & \texttt{Rec}\ [ ( \mathcal{L},\ expr ) ]
    \end{split}
  \end{equation*}
  Where:
  \begin{tabular}[t]{l @{ $=$ } l}
    $\mathcal{S}$ & string literals    \\
    $\mathcal{N}$ & numeric literals   \\
    $\mathcal{L}$ & local identifiers  \\
    $\mathcal{G}$ & global identifiers
  \end{tabular}

@@

We decided to focus our theory exploration research on Haskell as it has mature, state-of-the-art implementations (\qspec{} \citep{QuickSpec} and \hspec{} \citep{claessen2013automating}). This is evident from the fact that the state-of-the-art equivalent for Isabelle/HOL, the \textsc{Hipster} \citep{Hipster} system, is actually implemented by translating to Haskell and invoking \hspec{}.

Haskell is a programming language which is well-suited to research; indeed, this was a goal of the language's creators \citep{marlow2010haskell}. Like most members of the \emph{functional programming} paradigm, Haskell is essentially a variant of $\lambda$-calculus; in fact an intermediate representation of the \textsc{GHC} compiler, known as \emph{GHC Core}, is based on a variant of the polymorphic $\lambda$-calculus known as \fc{}. For simplicity, we will focus on this Core language rather than the relatively large and complex syntax of Haskell proper. For a full treatment of \fc{} and its use in GHC, see \citep[Appendix C]{sulzmann2007system}.

The sub-set of Core we consider is shown in figure \ref{coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major changes are to erase types and use a custom representation of names. There are also several other forms of literal (machine words of various sizes, individual characters, etc.) which we omit for brevity, as their treatment is similar to those of strings and numerals.

\subsection{QuickCheck}
\label{quickcheck}

@@

However, this is \emph{still} not enough, since this function will also accept fractional values\footnote{Since we only use generic numeric operations, the function will be polymorphic with a type of the form \hs{forall t. Num t => t -> t}, where \hs{Num t} constrains the type variable \hs{t} to be numeric. In fact, Haskell will infer extra constraints such as \hs{Eq t} since we have used \hs{==} in the unit tests.}, which will also cause it to diverge. Clearly, by choosing what to test we are biasing the test suite towards those cases we've already taken into account, whilst neglecting the problems we did not expect.

Haskell offers a partial solution to this problem in the form of \emph{property checking}. Tools such as \qcheck{} separate tests into three components: a \emph{property} to check, which unlike a unit test may contain \emph{free variables}; a source of values to instantiate these free variables; and a stopping criterion. Here is how we might restate our unit tests as properties:

@@

The ability to state \emph{universal} properties in this way avoids some of the bias we encountered with unit tests. In the \hs{factorial} example, this manifests in two ways:

\begin{itemize}
  \item \qcheck{} cannot test polymorphic functions; they must be \emph{monomorphised} first (instantiated to a particular concrete type). This is a technical limitation, since \qcheck{} must know which type of values to generate, but in our example it would bring the issue with fractional values to our attention.

@@

@@

\begin{figure}
  \begin{equation*}
    \begin{split}
      names\       =\ & \{ N_1, \dots, N_n \}                                                                  \\
      types\       =\ & \{ T_1, \dots, T_m \} \cup \{a \rightarrow b \mid a \in types \wedge b \in types \} \\
      exprs\       =\ & names \cup \{ f\ x \mid f \in exprs \wedge x \in exprs \wedge f : a \rightarrow b \wedge x : a \} \\
      classes\     =\ & \{c \mid \forall x y \in c, x \in exprs \wedge y \in exprs \wedge x =_{QC} y \} \\
      conjectures\ =\ & \bigcup_{c \in classes}\{ \ulcorner a = b \urcorner \mid a \in c \wedge b \in c \}
    \end{split}
  \end{equation*}

  \caption{General model of \qspec{} conjecture generation, given a signature $\Sigma$ containing names $N_1$ to $N_n$ and types $T_1$ to $T_m$. We use $a =_{QC} b$ to denote that $a$ and $b$ are indistinguishable by \qcheck{}, and use $\ulcorner \dots \urcorner$ for quasiquotation.}
  \label{qspecmodel}
\end{figure}

In this work we focus on the problem of \emph{(automated) theory exploration}, which includes the ability to \emph{generate} conjectures about code, to \emph{prove} those conjectures, and hence output \emph{novel} theorems without guidance from the user. In \citep{warburtonscaling} we identify the method of conjecture generation as a key characteristic of any theory exploration system, and in particular lament the existing reliance on brute force methods.

We focus on \qspec{} \citep{QuickSpec}, which discovers equations about Haskell code. These are found through the process modelled in figure \ref{qspecmodel}: given a typed signature $\Sigma$, \qspec{} enumerates all type-correct combinations of terms from $\Sigma$ up to some depth; it then groups them into equivalence classes using the \qcheck{} counterexample finder, then conjectures equations relating the members of these classes.

The resulting set of $conjectures$ can be used in several ways: it can be simplified for direct presentation to the user, sent to a more rigorous system like \hspec{} and \textsc{Hipster} for proving, or even serve as a background theory for an automated theorem prover \citep{claessen2013automating}.

\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.

In fact, there are similarities between the way a TE system like \qspec{} can generalise from checking \emph{particular} properties to \emph{inventing} new ones, and the way counterexample finders like \qcheck{} can generalise from testing \emph{particular} expressions to \emph{inventing} expressions to test. One of our aims is to understand the implications of this generalisation, the lessons that each can learn from the other's approach to term generation, and the consequences for testing and QA in general.

We attempt to mitigate the cost of running \qspec{}, by providing a machine learning layer analogous to that of premise selection in theorem proving. Of particular concern is the set of expressions $exprs$, which grows exponentially as the size of the expressions increases, and hence slows down the rest of the algorithm. Although complete, the current enumeration approach is wasteful: many combinations are unlikely to appear in theorems, which requires careful choice by the user of what to include in the signature. The key idea of premise selection is to make such choices automatically, by only including those expressions which are considered \emph{relevant} to the problem.

\subsection{Clustering and Feature Extraction}
\label{featureextraction}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{relevance}. Premise selection is a practical tool, used in production systems like the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} to compare expressions against. Instead, we are interested in relationships between \emph{any} components of a theory, and hence we must consider the relevance of \emph{everything} to \emph{everything else}. A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{clustering} for more information.

Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller sets more amenable to brute force enumeration.

Both of these approaches, premise selection and clustering, use \emph{machine learning} (ML) algorithms to analyse (logical or software) expressions, and hence rely on \emph{feature extraction} to transform the data into a suitable representation and remove irrelevant details. This has two benefits:

\begin{itemize}
  \item \emph{Feature vectors} (ordered sets of features) are chosen to be more compact than the data they're extracted from: feature extraction is \emph{lossy compression}. This reduces the size of the machine learning problem, improving efficiency (e.g. running time).
  \item We avoid learning irrelevant details, such as the text encoding system used, improving \emph{data} efficiency (the number of samples required to spot a pattern).
\end{itemize}

Another benefit of feature extraction is to \emph{normalise} the input data to a fixed-size representation. Many ML algorithms only work with inputs of a uniform size; feature extraction allows us to use these algorithms in domains where the size of each input is not known, may vary or may even be unbounded.

As an example, say we want to learn relationships between the following program fragments:

\begin{haskell}
data Maybe a = Nothing | Just a

data Either a b = Left a | Right b
\end{haskell}

We might hope our algorithm discovers relationships like:

\begin{itemize}
  \item Both are valid Haskell code.
  \item Both describe datatypes.
  \item Both datatypes have two constructors.
  \item \hs{Either} is a generalisation of \hs{Maybe} (we can define \hs{Maybe a = Either () a} and \hs{Nothing = Left ()}).
  \item There is a symmetry in \hs{Either}: \hs{Either a b} is equivalent to \hs{Either b a} if we swap occurences of \hs{Left} and \hs{Right}.
  \item It is trivial to satisfy \hs{Maybe} (using \hs{Nothing}).
  \item It is not trivial to satisfy \hs{Either}; we require an \hs{a} or a \hs{b}.
\end{itemize}

However, this is too optimistic. Without our domain-knowledge of Haskell, an ML algorithm cannot impose any structure on these fragments, and will treat them as strings of bits. Our high-level hopes are obscured by low-level details: the desirable patterns of Haskell types are mixed with undesirable patterns of ASCII bytes, of letter frequency in English words, and so on.

In theory we could throw more computing resources and data at a problem, but available hardware and corpora are always limited. Instead, feature extraction lets us narrow the ML problem to what we, with our domain knowlege, consider important.

There is no \emph{fundamental} difference between raw representations and features: the identity function is a valid feature extractor. Likewise, there is no crisp distinction between feature extraction and machine learning: a sufficiently-powerful learner doesn't require feature extraction, and a sufficiently-powerful feature extractor doesn't require any learning! \footnote{Consider a classification problem, to assign a label $l \in L$ to each input. If we only extract a single feature $f \in L$, we have solved the classification problem without using a separate learning step.}

Rather, the terms are distinguished for purely \emph{practical} reasons: by separating feature extraction from learning, we can distinguish straightforward, fast data transformation (feature extraction) from complex, slow statistical analysis (learning). This allows for modularity, separation of concerns, and in particular allows ``off-the-shelf'' ML to be re-used across a variety of different domains.

In our case, we will use a novel feature extraction algorithm, described in \S \ref{recurrentclustering}, to transform expressions in Haskell Core into a fixed size representation, suitable for clustering via the standard \emph{k-means} algorithm.

### 2015-11-17 TransferReport/contributions.tex
Updates

\section{Contributions}
\label{current}

\iffalse
TODO
Good analysis of differences with ML4PG/ACL2ML
\fi

@@

\subsubsection{Algorithm}

\begin{figure}
  \begin{haskell}
    type Env = [Local]
    type Db  = [(Global, Int)]

    data RoseTree a = Rose a [RoseTree a]

    exprTree :: Env -> Db -> Expr -> RoseTree Int
    exprTree env db e = case e of
      Var (Global i) -> Rose (fromMaybe sUNKNOWN (lookup db  i)) []
      Var (Local  i) -> Rose (fromMaybe sUNKNOWN (lookup env i)) []
      Lit (LitNum _) -> Rose (sLITNUM) []
      Lit (LitStr _) -> Rose (sLITSTR) []
      App  e1 e2     -> Rose (sAPP) [exprTree    env  db e1,
                                     exprTree    env  db e2]
      Lam  i  e      -> Rose (sLAM) [exprTree (i:env) db e1]
      Let  bs e      -> Rose (sLET)
      Case e i alts  -> Rose (sCASE) (exprTree env db e : map (altTree (i:env) db) alts)

    altTree :: Env -> Db -> Alt -> Rose Int
    altTree env db alt = case alt of
      (DataAlt _, vs, e) -> exprTree (vs ++ env) db e
      (LitAlt  _, _,  e) -> exprTree env db e
      (Default,   _,  e) -> exprTree env db e

    closedTree :: Db -> Core -> RoseTree Int
    closedTree = exprTree []
  \end{haskell}
  \caption{Feature extraction for Haskell Core, using recurrent clustering an k-means.}
  \label{featureextractionalgorithm}
\end{figure}

@@

\begin{itemize}
  \item Since they are comparatively rare, we ignore the particular value of each literal.
  \item Since we are currently focused on expressions, we ignore particular data constructors and types.
  \item If we cannot find a particular identifier, we use the sentinel value \hs{sUNKNOWN}. In practice, this only occurs for mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
  \item
\end{itemize}

@@

\iffalse
TODO
Replacing mutually-recursive calls (i.e. unknown global references) by the \hs{sUNKNOWN} sentinel, has the effect of denotes an has a similar effect to a consistently
\fi

@@

\begin{itemize}
  \item We cannot distinguish between data constructors, other than using exact equality.
  \item The arity of Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, there may be spurious after erasure rather than the Our Core expressions may contain type-level operation
\end{itemize}

effectively a collection contains  As a sub-set of Common Lisp, ACL2 is an untyped language; whilst Coq makes heavy use of types as first-class values.  Haskell differs from both (the ACL2 sub-set of) Common Lisp and Coq (by ML4PG) analyses an untyped language, Haskell effectively lives between the two extremes of untyped programming (as analysed by ACL2(ml)) and dependently typed programming (Coq (which is dependently typed) a sense, Haskell ACL2(ml)

>>>>

\section{Contributions}
\label{contributions}

@@

\subsubsection{Algorithm}
\label{algorithm}

\begin{figure}
  \begin{lstlisting}[language=Haskell, xleftmargin=.1\textwidth, xrightmargin=.1\textwidth]
    data RoseTree a = Node a [RoseTree a]

    eRT :: [Local] -> [[Global]] -> Expr -> RoseTree Feature
    eRT env db e = case e of
      Var (Global i) -> Node (lookupG db  i) []
      Var (Local  i) -> Node (lookupL env i) []
      Lit (LitNum _) -> Node sLITNUM []
      Lit (LitStr _) -> Node sLITSTR []
      Lam  i  e      -> Node sLAM [eRT (i:env) db e]
      App  e1 e2     -> Node sAPP [eRT    env  db e1,
                                   eRT    env  db e2]
      Let  bs e      -> Node sLET (map (bRT (ids bs:env) db) bs ++
                                        eRT (ids bs:env) db  e
      Case e i alts  -> Node sCASE (eRT    env  db  e :
                               map (aRT (i:env) db) alts)

    aRT :: [Local] -> [[Global]] -> Alt -> RoseTree Feature
    aRT env db alt = case alt of
      (DataAlt _, vs, e) -> eRT (vs ++ env) db e
      (LitAlt  _, _,  e) -> eRT env db e
      (Default,   _,  e) -> eRT env db e

    bRT :: [Local] -> [[Global]] -> Bind -> RoseTree Feature
    bRT env db b = case b of
      NonRec i e -> eRT env db e
      Rec es     -> Node sREC (map (eRT env db . snd) es)

    level :: Int -> RoseTree Feature -> [Feature]
    level 0 (Node x xs) = [x]
    level n (Node x xs) = concatMap (level (n-1)) xs

    rt :: [[Global]] -> Expr -> [Feature]
    rt db e = concat (pad cols (map (`level` eRT [] db e) [0..rows]))

    recurrentCluster :: [[(Global, Expr)]] -> [[Global]]
    recurrentCluster = go ([],[])
      where go (fs, db) []       = db
            go (fs, db) (es:ess) = let fs' = fs ++ map (extract db) es
                                       db' = kMeans fs'
                                    in go (fs', db') ess
            extract db (i, e) = (i, rt db e)
  \end{lstlisting}
  \caption{Feature extraction for Haskell Core. We send named expressions to \hs{recurrentCluster}, topologically sorted: if an element contains multiple \hs{(Global, Expr)} pairs, they are mutually-recursive. Syntax trees are first converted into \hs{RoseTree}s of \hs{Feature}s, these trees are converted to matrices, then flattened into feature vectors, then \hs{kMeans} clusters the feature vectors. The result is a list of clusters, naming their elements. More details are given in \S \ref{algorithm}.}
  \label{featureextractionalgorithm}
\end{figure}

@@

\begin{itemize}
  \item Since they are comparatively rare, we ignore the particular value of each literal.
  \item Since we are currently focused on expressions, we ignore particular data constructors and types.
  \item The \hs{lookupL} and \hs{lookupG} functions return the index containing their argument, if found. In the case of \hs{lookupL}, this acts as a de Bruijn index to give alpha-equivalent terms equal feature vectors. For \hs{lookupG}, this is the ID of the cluster it appears in; this ensures that references to similar expressions result in similar features.
  \item \hs{lookupL} and \hs{lookupG} return a sentinel value \hs{sUNKNOWN} if the given ID is not found. In practice, this only occurs for global, mutually-recursive references, since the toplogical sort will populate the \hs{db} with all other global identifiers, and local identifiers which don't occur in the environment would be rejected as invalid Haskell before the translation to Core.
  \item The \hs{Feature} type is abstract, but in practice it will be \hs{Float}.
  \item Our algorithm contains several parameters, including \hs{rows} and \hs{cols} which determine how to truncate the matrices (defaults are 30). The \hs{kMeans} function also contains a parameter for the cluster number; we set this as $\sqrt{n}$ where $n$ is the number of feature vectors being clustered.
\end{itemize}

@@

@@

\begin{itemize}
  \item Although they are values, we cannot distinguish between data constructors, other than using exact equality.
  \item Since Core uses a single \texttt{Lam} abstraction for both value- and type-level parameters, we cannot completely erase type-level parameters unless they are fully applied. This can cause a function's Core arity to be greater than its Haskell arity.
\end{itemize}

There is certainly promise in including types in our analysis, by pairing every term with its type as in ML4PG. This will allow fine-grained distinction of expressions which are otherwise identical, especially data constructors.
